Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-11-09 doc/manuscript.tex
updates to model and results

Genomic surveys of natural systems suggest that historical admixture among
diverged populations or closely related taxa commonly occors
\citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
is potentially widespread in primate
\citep{tung2017contribution,sorensen2023genome} and hominin
\citep{wolf2018outstanding,peter2020100} evolution. Admixture may therefore be
an important driver of phenotypic and molecular variation and shape the genetic
basis of complex traits. In humans, archaic introgression involving
Neanderthals and Denisovans has attracted considerable attention, with efforts
to describe the historical processes leading to observed distributions of
introgressed DNA in present-day populations
\citep{prufer2014complete,villanea2019multiple,chen2020identifying} and its
contribution to quantitative traits
\citep{sankararaman2016combined,wei2023lingering}.

@@

There is growing genetic evidence that early \emph{H. sapiens} reciprocally
contributed to Neanderthal genomes
\citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
human-to-Neanderthal introgression would have occured tens or hundreds of
thousands of years prior to Neanderthal introgression in humans during modern
human global dispersal around 60 ka. This is supported by ``near modern''
\emph{H. sapiens} outside of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary, up to $\sim5\%$ of later Neanderthal genomes may have been
contributed through introgression from humans to Neanderthals
\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
fewer deleterious alleles due to their larger long-term effective population
size, human-introgressed DNA would have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports such a model of post-admixture selection in the Neanderthal
lineage \citep{posth2017deeply,petr2020evolutionary}.

@@

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:one-pop}A). When the mean phenotype of the population is close
to the phenotypic optimum, classical models predict that alleles contributing
to such a trait are subject to underdominant selection, i.e., selection against
the minor allele at a locus \citep{robertson1956effect}. This has proven to be
a useful model for understanding allele frequencies of variants contributing to
a trait under stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
        \aprcomment{$w$ and $f$ need to match text}
        \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
    }
    \label{fig:one-pop}
\end{figure}

In diverged populations, the genomic architecture of a trait under stabilizing
selection has a higher rate of turnover compared to neutrally evolving loci
\citep{yair2022population}. We might therefore expect an accumulation of fixed
differences in the \emph{H. sapiens} and Neanderthal lineages at loci
contributing to a trait, even when the mean phenotype in each population is
close to the same trait optimum. When a derived allele at high frequency in one
population is introduced to another population in which is was previously
absent, it will be at low frequency and subsequently selected against.
Likewise, if the ancestral allele is reintroduced to a population fixed for a
derived allele, the ancestral allele will be at low frequency and will be
selected against. In either case we should expect selection against the
introgressed allele, whether ancestral or derived and regardless the historical
relative sizes of the populations involved. This prediction contrasts with the
population-genetics ``load'' model, in which haplotypes with fewer deleterious
variants (such as those from the population with larger historical size) are
favored after introgression in either direction.

@@

We consider a polygenic trait such that an individual's (additive) genetic
value is the sum over all effects of alleles in their genome, so that for
individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
genotype at that locus (i.e., the number of alleles they carry at that locus).
For unlinked loci, the expected additive genetic variance is \(V_A = \sum_l
2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We ignore
dominance and epistatic effects, so that \(V_G = V_A\), assuming no linkage.
We further ingore environmental effects \citep{simons2018population}, so that
the phenotypic variance \(V_P=V_A\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$ (typically set to zero), and we assume a Gaussian fitness function so
that relative fitness is given by \(f(G_i | O, V_S) = \exp{(-(G_i - O)^2 / 2
V_S)}\). $V_S$ can be interpreted as the strength of selection on the trait,
where larger $V_S$ implies weaker selection. For a population with mean
phenotype at (or very close to) the optimum, the mean fitness of the population
(assuming a roughly normal distribution of phenotypic values in the population)
is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0, V_S) \mathcal{N}(0, V_G) dG
= \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus, as the genetic variance
increases, mean fitness among individuals in the population decreases.

\subsubsection*{Mutation rates and effect sizes}

If all alleles contribute equally to the trait
(with effect sizes \(\pm a\) occurring in equal proportion),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result
\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
Mutations are not expected to each have the same effect size $a$, but
rather will be drawn from some distribution. Throughout, we assume mutation
effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$.

\aprcomment{where does the $4\mu Ne V_M$ (Lande) result come from? point to fig 1}

@@

Figure~\ref{fig:one-pop}C shows that if $V_G$ is non-negligible compared to
$V_S$, ignoring $V_G$ and using $s=a^2/2V_S$ leads to lower estimates of
additive genetic variance. When mutation rates are large so that $V_G$ is not
small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of $V_G$
that closely match simulations assuming free recombination between loci
(Figure~\ref{fig:toy-admixture}C,D). Because $V_G$ can change over time, this
means that $s$ is no longer constant, but can change due to factors such as
non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:toy-admixture}
\end{figure}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{fig:toy-admixture}A,B).

@@

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is kept at $0$ in all branches, and the strength of selection
$V_S=1$ remains constant.

@@

We expect genetic variance to increase after introgression. The amount that
genetic variance increases depends on the allelic differences accumulated
between populations and the effects of those alleles. Assuming no linkage,
dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
proportion $f$ contributed by the source population (labeled 0) into the focal
population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
expression for $V_G$, and after some simple algebra (Appendix~AX), we can
express the expected genetic variance directly after admixture as \[V_G =
fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
p_1)^2$ is the squared difference in allele frequencies at a locus
\citep{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$.

In general, $F_2$ at a given locus will depend on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on the demography. In this case
\begin{align*}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align*}
where $V_M$ is the variance of effect sizes of new mutations (assuming the mean
is zero), and $L$ is the number of trait affecting loci.

For $a \not\approx 0$, the assumption of neutral evolution will lead to
overestimates of $F_2$ compared to underdominant selection. We can compute
expected $V_G$ both before and after admixture using the diffusion
approximation for the joint SFS with underdominance. Comparing to simulations
with free recombination between loci (Appendix~AY), we find that this provides
an excellent fit to average observed $V_G$ over time
(Figure~\ref{fig:toy-admixture}C,D). In general, admixture causes a sudden
increase in $V_G$ followed by a fairly rapid return to pre-admixture levels.

Modeling the dynamics of genetic variance using the SFS lets us examine
contributions to $V_G$ from different classes of mutations, such as those at
different frequencies or arising at different times, and how those
contributions change over time. Of interest is the contribution to $V_G$ from
mutations that were already segregating in the recipient population, those that
were introduced through introgression, and new mutations since the time of
admixture.

Paragraph on the following (point to appendix and supplement):
\begin{itemize}
    \item Benefit of SFS -- partition by frequency classes (prev. seg vs introgressed)
    \item VG from all both previously segregating and introgressed alleles decays,
        replaced by VG due to new mutations
    \item Dividing VG from introgressed alleles into introduced derived vs reintroduced
        ancestral alleles shows the effect of differences in population sizes
\end{itemize}

\subsection*{Recurrent introgression and non-constant population size}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/neanderthal_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:neand-to-human}
\end{figure}

\begin{itemize}
    \item In the demographic model more closely resembling inferred history
        connecting humans and Neanderthals (Figure~\ref{fig:neand-to-human}A),
        we see that demographic history and the distribution of effect sizes
        of new mutations impact $V_G$
    \item Additive genetic variance increases after introgression in both directions,
        which is expected to result in a period of decreased mean fitness
    \item Population contractions decrease $V_G$, as the increased rate of drift
        reduces diversity at loci underlying the trait
    \item TODO: compare to linked simulations
\end{itemize}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/human_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:human-to-neand}
\end{figure}

\begin{itemize}
    \item Because we know if a mutation was previously segregating in the population
        receiving migrants, we can partition variation by contributions from
        segregating vs introgressed variants (whether introduced derived mutations
        or reintroduced ancestral mutations)
    \item At low introgression proportions, as here, the majority of $V_G$ is still
        contributed by previously segregating mutations, which is then expected to
        be steadily replaced by new mutations (Figure 2). $V_G$ due to introgressed
        variation decays monotonically over time (Figure~\ref{fig:neand-to-human}C,D)
    \item While the total $V_G$ due to introgressed variants decays, we can also
        consider the average contribution of introgressed vs non-introgressed SNPs
        to $V_G$ ($h^2$ per SNP) -- at least for this demographic model and
        DES, the contribution per SNP of introgressed variants is initially lower
        than those of previously segregating variants. These contributions change
        over time, in a way that depends on both the genomic architecture underlying
        the trait and demography
    \item This dependence on both architecture and demography is apparent when
        comparing to $V_G$ contributed by introgressed and non-introgressed variants
        in the early human-to-Neanderthal admixture (Figure~\ref{fig:human-to-neand})
        -- here, again the contribution
        from introgressed variants to $V_G$ decays monotonically, but depending on
        the distribution of effect sizes, relative per-SNP $h^2$ changes over time
        (though in these cases introgressed $h^2$ per SNP remains lower than
        non-introgressed, when considering mutations at matched allele frequencies)
\end{itemize}

\subsection*{The effects of linkage}

\begin{itemize}
    \item So far, we have considered model and simulation without linkage -- in this
        case, the underdominance model provides an excellent approximation
    \item Including linkage between sites dramatically complicates things, making
        theorical or analytical progress difficult
    \item To explore the effects of linkage, we used individual-based simulations
        with a single large chromosome of size 1 Morgan, and varried the total
        mutation rate per haploid and the variance of effect sizes of new mutations
    \item We'll highlight two effects of linkage
        \begin{itemize}
            \item Deviations from expected $V_G$, as well as more complicated
                short-term dynamics after admixture
            \item The reduction in introgressed ancestry in the regions surrounding
                trait-affecting alleles
        \end{itemize}
\end{itemize}

\subsubsection*{Effects of linkage on expected genetic variance}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/model_comparison.SD_0.01.mu_0.001_0.025}
    \caption{}
    \label{fig:linkage-deviations}
\end{figure}

\begin{itemize}
    \item Stabilizing selection reduces $V_G$ by selecting against minor alleles
        (underdominance), but with linkage it can also reduce $V_G$ by selecting
        for haplotypes that carry opposite effect alleles that cancel out -- this
        creates LD between positive and negative effect alleles at linked loci,
        known as the Bulmer effect \citep{}
    \item Thus, $V_G$ in simulations with linkage deviate from predictions
        assuming underdominance alone
    \item The severity of this deviation depends on a number of factors: mean
        recombination rate between trait-affecting loci, the total mutation rate,
        the effect sizes (and distribution of effect sizes of new mutations)
    \item In general, lower mutation rates imply lower total $V_G$, lower polygenicity,
        and therefore higher recombination rates between trait-affecting alleles --
        we therefore expect this deviation to be exacerbated with higher mutation
        rates (all else being equal)
\end{itemize}

\subsubsection*{Reduction of introgressed ancestry surrounding an introgressed trait-affecting allele}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/linkage_predictions.pdf}
    \caption{
        \textbf{}
        \aprcomment{Linked allele frequency should always be minor, and labeled
        introgressed ancestry frequency}
    }
    \label{fig:linkage-pred}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/linkage_simulation.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:linkage-sim}
\end{figure}

@@

\section{Appendix}

\subsection{Additive genetic variance after admixture}

@@

\subsection{Dynamics of introgressed alleles}

\subsection{Reciprocal depletion of functional variation}

>>>>

Genomic surveys of natural systems suggest that historical admixture among
diverged populations or closely related taxa commonly occurs
\citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
is potentially widespread in primate
\citep{tung2017contribution,sorensen2023genome} and hominin
\citep{wolf2018outstanding,peter2020100} evolution. Admixture may therefore be
an important driver of phenotypic and molecular variation and shape the genetic
basis of complex traits. In humans, archaic introgression involving
Neanderthals and Denisovans has attracted considerable attention, with efforts
to describe the historical processes leading to observed distributions of
introgressed DNA in present-day populations
\citep{prufer2014complete,villanea2019multiple,chen2020identifying} and its
contribution to quantitative traits
\citep{sankararaman2016combined,wei2023lingering}.

@@

There is growing genetic evidence that early \emph{H. sapiens} reciprocally
contributed to Neanderthal genomes
\citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
human-to-Neanderthal introgression occured tens or hundreds of thousands of
years prior to Neanderthal introgression in humans during modern human global
dispersal around 60 ka. This is supported by ``near modern'' \emph{H. sapiens}
outside of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary, around $6\%$ of later Neanderthal genomes may have been
contributed through introgression from humans to Neanderthals
\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
fewer deleterious alleles due to their larger long-term effective population
size, human-introgressed DNA would have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports such a model of post-admixture selection in the Neanderthal
lineage \citep{posth2017deeply,petr2020evolutionary}.

@@

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is close
to the phenotypic optimum, classical models predict that alleles contributing
to such a trait are subject to underdominant selection, i.e., selection against
the minor allele at a locus \citep{robertson1956effect}. This has proven to be
a useful model for understanding genetic architectures of traits under
stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
        \aprcomment{$w$ and $f$ need to match text}
        \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
        \aprcomment{First figure should just be two left panels. Combine two
        right panels to figure 2.}
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutrally
evolving loci \citep{yair2022population}. We might therefore expect an
accumulation of fixed differences in the \emph{H. sapiens} and Neanderthal
lineages at loci contributing to such a trait, even when the mean phenotype in
each population remains close to the same trait optimum. When a derived allele
at high frequency in one population is introduced to another population in
which is was previously absent, it will be at low frequency and subsequently
selected against. Likewise, if the ancestral allele is reintroduced to a
population fixed for a derived allele, the ancestral allele will be at low
frequency and will be selected against. In either case we should expect
selection against the introgressed allele, whether ancestral or derived and
regardless the historical relative sizes of the populations involved. This
prediction contrasts with the population-genetics ``load'' model, in which
haplotypes with fewer deleterious variants (such as those from the population
with larger historical size) are favored after introgression in either
direction.

@@

We consider a polygenic trait for which an individual's (additive) genetic
value is the sum over all effects of alleles in their genome, so that for
individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
genotype at that locus (i.e., the number of alleles they carry at that locus).
With no linkage between trait-affecting loci, the expected additive genetic
variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
frequency at locus $l$. We ignore dominance and epistatic effects, so that
\(V_G = V_A\).  We further ingore environmental effects
\citep{simons2018population}, so that the phenotypic variance \(V_P=V_A\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$ (typally set to zero), and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $V_S$ can be interpreted as the strength
of selection on the trait, where larger $V_S$ implies weaker selection. For a
population with mean phenotype at (or very close to) the optimum, the mean
fitness of the population (assuming a roughly normal distribution of phenotypic
values in the population) is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0,
V_S) \mathcal{N}(0, V_G) dG = \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus,
as the genetic variance increases, mean fitness among individuals in the
population decreases.

\subsubsection*{Mutation rates, effect sizes and genetic variance}

If all alleles contribute equally to the trait
(with effect sizes \(\pm a\) occurring in equal proportion),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result for
steady-state additive genetic variance,
\[V_G \approx 4\mu V_S,\] provided \(V_G \ll V_S\).

Mutations are not expected to each have the same effect size $a$, but rather
follow some distribution. Throughout, we assume effect sizes are drawn from a
normal distribution with mean 0 and given variance $V_M$. When population sizes
or effect sizes are small, drift dominates the dynamics of $V_G$, and at steady
state \citet{lande1976natural} found \[V_G \approx 4 N_e \mu V_M.\]
Interpolating between the selection- and drift-dominated regimes,
\begin{align}\label{eq:SHC}
    V_G & \approx \frac{4 \mu V_S}{1 + \frac{VS}{N_e V_M}}.
\end{align}
This, the ``stochastic
house-of-Cards'' (SHC) approximation (Figure~\ref{fig:stab-sel}B), was given by
\citet{burger1989much} and is discussed in detail in \citet[][Ch.
28]{walsh2018evolution}.

@@

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=a^2/2V_S$ leads to lower estimates of additive genetic variance
(Figure~\ref{fig:toy-admixture}B). When mutation rates are large so that $V_G$
is not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
$V_G$ that closely match simulations assuming free recombination between loci
(Figure~\ref{fig:toy-admixture}X,Y). Because $V_G$ can change over time, this
means that $s$ is no longer constant, but can change due to factors such as
non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{Predicting the genetic variance of a trait under stabilizing
        seletion and non-steady state demography.}
        \aprcomment{Make it in generations?}
        \aprcomment{Right two panels should go to supplement, move single-pop
        with size change from figure 1 to here.}
    }
    \label{fig:toy-admixture}
\end{figure}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{fig:toy-admixture}X,Y).

@@

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is kept at $0$ in all branches, so that no optimum shift occurs,
and the strength of selection $V_S=1$ remains constant.

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/neanderthal_admixture.pdf}
    \caption{
        \textbf{}
        \aprcomment{Show total/introgressed/non-introgressed in both directions
        with both $\sigma_M$.}
    }
    \label{fig:neand-to-human}
\end{figure}

@@

We expect genetic variance to increase after introgression. The amount that
genetic variance increases depends on the allelic differences accumulated
between populations and the effects of those alleles. Assuming no linkage,
dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
proportion $f$ contributed by the source population (labeled 0) into the focal
population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
expression for $V_G$ and after some simple algebra (Appendix,
section~\ref{sec:VG-admixture}), we can express the expected genetic variance
directly after admixture as \[V_G = fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l
F_{2,l} a_l^2,\] where $F_2 = (p_0 - p_1)^2$ is the squared difference in
allele frequencies at a locus \citep{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$.

$F_2$ at a given locus depends on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on the demography. In this case,
\begin{align}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    \nonumber
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align}
where $L$ is the number of trait affecting loci.

\subsubsection*{Predicted $V_G$ from the SFS}

For $a \not\approx 0$, the assumption of neutral evolution will lead to
overestimates of $F_2$ compared to underdominant selection. We can compute
expected $V_G$ both before and after admixture using the diffusion
approximation for the joint SFS with underdominance. Comparing to simulations
with free recombination between loci (Appendix,
section~\ref{sec:VG-admixture}), we find that this provides an excellent fit to
average observed $V_G$ over time (Figure~\ref{fig:toy-admixture}C,D). In
general, admixture causes a sudden increase in $V_G$ followed by a fairly rapid
return to pre-admixture levels.

Modeling the dynamics of genetic variance using the SFS lets us examine
contributions to $V_G$ from different classes of mutations, such as those at
different frequencies or arising at different times, and how those
contributions change over time. Of interest is the contribution to $V_G$ from
alleles that were already segregating in the recipient population, those that
were introduced through introgression, and new mutations since the time of
admixture (Figure~\ref{fig:toy-admixture}). In many scenarios, the initial
divergence of the two populations will have occurred long enough before the
admixture event, so that the mutations underlying $V_G$ are largely unique in
each population. After mixing, previously segregating and introgressed alleles
go to fixation or loss, and the variance of the trait is increasingly due to
new mutations.

Introgressed variation can initially make up a considerable portion of $V_G$,
with those alleles being either newly introduced derived alleles or
reintroduced ancestral alleles. The relative sizes of the two populations
impacts the number of each, as derived alleles will accumulate more readily in
a population with smaller effective size. Nonetheless, the increase in $V_G$
can be fairly similar in either direction of introgrssion, as the term
\(2f(1-f)\sum_l F_{2,l} a_l^2\) contributes in either case and can be much
larger than $f V_G$ from the source population.

\subsubsection*{Complex demography and partitioning heritability by origin of alleles}

We used a historical model more closely resembling inferred human-Neanderthal
history (Figure~\ref{fig:neand-to-human}A) to explore the effects of population
size changes and reciprocal admixture on the additive genetic architecture of
traits under stabilizing selection. As expected
(Figure~\ref{fig:toy-admixture}), population contractions decrease $V_G$ as the
increased rate of drift reduces allelic diversity at trait-affecting loci, and
introgression increases $V_G$ in both directions of gene flow.

Because the genetic architectures considered here are purely additive, we can
track mutations in an admixed populations by whether they were previously
segregating, fixed or lost in either parental populations or if they arose as
new mutations since the time of admixture (Methods Section X). This allows us
to partition $V_G$ by contributions from these different sets of mutations. At
low introgression proportions, as modeled here, the majority of $V_G$ is still
contributed by previously segregating, non-introgressed mutations (Figure XAB).
$V_G$ due to existing mutations decays monotonically over time and is fairly
rapidly replace by new mutations (Figure XCD).

The average contribution of introgressed vs. non-introgressed SNPs to $V_G$
(i.e., $h^2$ per SNP) can similarly by tracked over time. For the
human-Neanderthal-like demographic model and genetic architectures considered
here, the contribution per SNP of introgressed variants is initally lower than
that of non-introgressed variants. These contributions change over time,
depending on mutational variance (Figure YA vs B), as well as demography
(comparing the two directions of introgression, Figure YC D). When weighting
$h^2$ per SNP of introgressed SNPs by matching to allele frequencies of
non-introgressed variants (Methods), relative contributions depend sensitively
on evolutionary parameters and the time since admixture.

\aprcomment{In discussion, point out that this makes interpreting results, such
as \cite{wei2023lingering}, challenging.}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/human_admixture.pdf}
    \caption{
        \textbf{}
        \aprcomment{Put $h^2$ per SNP into its own figure.}
    }
    \label{fig:human-to-neand}
\end{figure}

\subsection*{Expected genetic variance with linkage}

In the preceding sections, we find that approximating the dynamics of
trait-affecting alleles using an underdominance model
\citep{robertson1956effect} provides an excellent approximation of $V_G$ in
complex demographic scenarios. However, this relies on an absence of linkage
between trait-affecting alleles. The inclusion of linkage can lead to
noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} slightly
overestimates $V_G$ at steady state
\citep{burger1989much,burger1994distribution,walsh2018evolution}.

To investigate the effects of linkage, we used chromosome-scale
individual-based simulations. By varying the mutation rate and the variance of
effect sizes of new mutations, we include scenarios ranging from low to high
polygenicity and from weak to strong selection on individual alleles. In this
and the following sections, we will highlight two effects of linkage. First,
we observe deviations of $V_G$ from expectations without linkage, which can be
large for highly polygenic traits. Second, selection on introgressed
trait-affecting alleles results in a reduction of introgressed ancestry in
surrounding regions.

With linkage between two or more alleles affecting a trait under stabilizing
selection, linkage disequilibrium (LD) can develop between alleles. Classicaly,
stabilizing selection will cause mutations of opposite-signed effects to be in
coupling LD and those of same-signed effects in repulsion LD
\citep{bulmer1971effect}. This has the effect of reducing $V_G$. For lower
mutational variances ($2N_e V_M\approx1$, Figure~\ref{fig:low_VM}), we observe
this phenomenon. With low mutational input, and thus low polygenicity, $V_G$ in
individual simulations closely matches expectations without linkage. As we
increase the mutation rate, $V_G$ is reduced relative to those expectations.
However, when the mutational variance is much larger ($2Ne V_m \approx 50$), we
see the reverse trend (Figure~\ref{fig:high_VM}). At low mutation rate, there
is a close match between observed $V_G$ and unlinked expectations, although
simulated values are slightly higher. As the mutation rate increases, $V_G$
increases to be relatively much larger that expectations, rather than smaller.

The strength of the deviation of $V_G$ between models with and without linkage
depends on a number of factors. The total mutation rate affects not only the
polygenicity, with higher mutation rates causing higher number of segregating
alleles, but also the mean recombination rate between those alleles, as they
will be more or less densely distributed in the genome. The distribution of
effect sizes plays an important role, as seen by the opposite trends in $V_G$
relative to unlinked predictions, which are most apparent with high mutation
rates. There appear to be complex dynamics involving the Bulmer Effect
\citep{bulmer1971effect} and interference between selected alleles
\citep{hill1966effect}, which warrants careful investigation.

\subsubsection*{Introgressed ancestry is reduced around introduced trait-affecting alleles}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/linkage_predictions.pdf}
    \caption{
        \textbf{}
        \aprcomment{Linked allele frequency should always be minor, and labeled
        introgressed ancestry frequency}
        \aprcomment{Combine with next figure}
    }
    \label{fig:linkage-pred}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/linkage_simulation.pdf}
    \caption{
        \textbf{}
        \aprcomment{Combine with previous figure}
    }
    \label{fig:linkage-sim}
\end{figure}

\subsection*{Introgressed ancestry deserts are shared under reciprocal introgression}

@@

\appendix

\section{Additive genetic variance after admixture}\label{sec:VG-admixture}

@@

\section{Dynamics of introgressed alleles}

\section{Reciprocal depletion of functional variation}
